---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vip

<!-- badges: start -->
<!-- badges: end -->

**Vaccine Immunity Propagation**

This repository contains functions to run a demographic model of
vaccine exposure over time, tracking the vaccine-derived immunity by
age through time, following implementation of one or more vaccination
activities that can target the population as a whole or selectively
target specific age groups.

## Installation

You can install the development version of vip from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mrc-ide/vip")
```

## Example

This package defines an S3 class "vip_population" which is a dataframe
with at least the columns year, age, cohort (the year of birth for the
individuals tracked in this row), and immunity (as a proportion of
this cohort). Year, age and cohort are tracked as annual time steps
and age groups.

A basic, totally naive population dataframe can be setup using the
function "vip_population":

```{r example}
library(vip)

pop <- vip_population(year_min = 2000, year_max = 2010,
                        age_min = 0, age_max = 10)

dim(pop)
head(pop)

```
This dataframe has 121 columns (11 age groups 0 - 10 x 11 years 2000 - 2020). 

Next, we read in a file containing some vaccination activites into an object of the class 'vip_vacc_activities'. 

The first is a campaign targeting all age groups of our
dummy population, which took place 5 years before the dawn of time. As
immunity is assumed not to wane, the effects of this are still there,
and therefore the functions here keep track of this. The remaining
vaccination activities are routine vaccination of infants (restricted
to age 0) with an increasing population of the target cohort to be
vaccinated.

We now apply these vaccination activities sequentially to the
population using the function 'apply_vacc'.

```{r apply_vacc}
library(dplyr) ## needed for the pipe %>%

vaccs <- read_vacc_activities("inst/exdata/vacc_activities.csv")
vaccs

pop <- apply_vaccs(pop, vaccs)

```

The resulting vaccine-derived immunity of the population can be visualised with
the function "plot_population()". This is based on ggplot2, and the
returned graph object can be further modified - here in order to
achieve tick marks that suit the data better.

```{r plot_population}
library(ggplot2) ## needed to amend the plot returned from plot_population()

g <- plot_population(pop)
g + scale_x_continuous(breaks = seq(2000, 2010, by = 2)) +
    scale_y_continuous(breaks = seq(0, 10, by = 2))

```

In the top left corner it shows the immunity remaining in the older
age groups from the initial campaign, while the infant vaccination
from 2005 onwards results in immuity in the lower right corner - and
the increasing coverage highlights how cohorts age through time and
therefore move through the plot in a diagonal fashion.


<!-- You'll still need to render `README.Rmd` regularly, to keep -->
<!-- `README.md` up-to-date. `devtools::build_readme()` is handy for -->
<!-- this. You could also use GitHub Actions to re-render `README.Rmd` -->
<!-- every time you push. An example workflow can be found here: -->
<!-- <https://github.com/r-lib/actions/tree/v1/examples>. -->

<!-- You can also embed plots, for example: -->

<!-- ```{r pressure, echo = FALSE} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- In that case, don't forget to commit and push the resulting figure -->
<!-- files, so they display on GitHub and CRAN. -->
