---
title: "popim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{popim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(popim)
```

The intention of the package `popim` is to provide tools to easily
evaluate and track the POPulation IMmunity of an age-structured
population that results from vaccination. Given that vaccination has
long lasting effects, this requires tracking the status of the
population through time.

## Data structures

To this end the package defines two S3 classes to store data in an
appropriate format. Both classes are dataframes, but have additional
requirements such as certain columns that must be present, and the
format and range of these columns.

### S3 class `popim_population`

The class `popim_population` is designed to hold information on a
population of interest.

#### The dataframe

The dataframe holds information on the population size and immunity
status of the population through time. The population is disaggregated
into 1-year age groups, and potentially several spatial regions. Each
row refers to a particular (1 year) birth cohort in a given location
and year.

The mandatory columns are:

* `region`: The population of interest may be spatially
  disaggregated into separate geographical regions (e.g., countries or
  subnational administrative regions). This column holds a string that
  identifies which region each entry refers to.
* `year`: The population is tracked through time using annual time
  steps.
* `age`: The age-structure of the population is tracked in annual age
  groups.
* `cohort`: This column is redundant as it equals `year` - `age`, and
  therefore refers to the year in which this cohort was born. It is
  included for ease of handling.
* `immunity`: Gives the proportion of each cohort that is immune due
  to past vaccination.
* `pop_size`: Number of people in any given age group and year. To
  facilitate using common units such as thousands this is not limited
  to integer values. It is however up to the user to ensure that such
  units are used consistently.
  
| column     | type      | range          |
|------------|-----------|----------------|
| `region`   | character |                |
| `year`     | integer   |                |
| `age`      | integer   | non-negative   |
| `cohort`   | integer   | `year` - `age` |
| `immunity` | numeric   | [0, 1]         |
| `pop_size` | numeric   | non-negative   |



#### Class attributes

In addition to the attributes that every dataframe has (`names`,
`class`, `row.names`), the objects of the `popim_population` class
also retain the input parameters of the basic constructer
`popim_population()`:

* `region`: a character vector of all regions in this object,
* `year_min` and `year_max`: the range of years, and
* `age_min` and `age_max`: the age range covered in this population.


#### Create, modify and validate objects of this class

create: [popim_population()], [`read_popim_pop()`], `as_popim_pop()`

modify: [apply_vacc()]

validate: [is_population()]




### S3 class `popim_vacc_activities`

The class `popim_vacc_activities` is designed to hold information on
vaccination activities that have occurred or are planned in the
population of interest. Each row of the dataframe relates to one
vaccination activity.

#### The dataframe

The mandatory columns are:

* `region`: The region in which the vaccination activity takes
  place. While this is a free character field, when applied to a
  population, this must match a region that occurs in the population.
* `year`: year during which the vaccination activity takes place.
* `age_first`: youngest age group to be targeted.
* `age_last`: oldest age group to be targeted.
* `coverage`: proportion of the target population to be immunised.
* `doses`: number of doses used in the vaccination activity.
* `targeting`: determines how doses are allocated when there is
  pre-existing immunity in the population.

*to do: details of coverage/doses, include assumptions of non-waning,
100% effectiveness, no wastage.*

*to do: explain targeting.*




| column      | type      | range             |
|-------------|-----------|-------------------|
| `region`    | character |                   |
| `year`      | integer   |                   |
| `age_first` | integer   | non-negative      |
| `age_last`  | integer   | $\ge$ `age_first` |
| `coverage`  | numeric   | [0, 1]            |
| `doses`     | numeric   | non-negative      |
| `targeting` | character | `"random"`, `"correlated"`, `"targeted"` |

#### Class attributes

As this is a subclass of dataframe, it has the same attributes as a
dataframe (`names`, `class` and `row_names`), and no additional
attributes.

#### Create and validate objects of this class

create: [popim_vacc_activities()], [read_vacc_activities()],
[as_vacc_activities()], [vacc_from_immunity()]

modify: [complete_vacc_activities()]

validate: [is_vacc_activities()]

## Primary functionality

Once objects holding data on the population and vaccination activities
have been set up, the primary functionality provided by `popim` is to
apply the vaccination activities to the population to evaluate the
population immunity by age over time that results from the given
vaccination activities, which is done with the function
[apply_vacc()].


## Visualisation
