---
title: "popim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{popim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

```{r setup}
library(popim)
```

The intention of the package `popim` is to provide tools to easily
evaluate and track the POPulation IMmunity of an age-structured
population that results from vaccination. Given that vaccination has
long lasting effects, this requires tracking the status of the
population through time.

## Data structures

To this end the package defines two S3 classes to store data in an
appropriate format. Both classes are dataframes, but have additional
requirements such as certain columns that must be present, and the
format and range of these columns.

### S3 class `popim_population`

The class `popim_population` is designed to hold information on a
population of interest.

#### The dataframe

The dataframe holds information on the population size and immunity
status of the population through time. The population is disaggregated
into 1-year age groups, and potentially several spatial regions. Each
row refers to a particular (1 year) birth cohort in a given location
and year.

The mandatory columns are:

* `region`: The population of interest may be spatially
  disaggregated into separate geographical regions (e.g., countries or
  subnational administrative regions). This column holds a string that
  identifies which region each entry refers to.
* `year`: The population is tracked through time using annual time
  steps.
* `age`: The age-structure of the population is tracked in annual age
  groups.
* `cohort`: This column is redundant as it equals `year` - `age`, and
  therefore refers to the year in which this cohort was born. It is
  included for ease of handling.
* `immunity`: Gives the proportion of each cohort that is immune due
  to past vaccination.
* `pop_size`: Number of people in any given age group and year. To
  facilitate using common units such as thousands this is not limited
  to integer values. It is however up to the user to ensure that such
  units are used consistently.
  
| column     | type      | range          |
|------------|-----------|----------------|
| `region`   | character |                |
| `year`     | integer   |                |
| `age`      | integer   | non-negative   |
| `cohort`   | integer   | `year` - `age` |
| `immunity` | numeric   | [0, 1]         |
| `pop_size` | numeric   | non-negative   |



#### Class attributes

In addition to the attributes that every dataframe has (`names`,
`class`, `row.names`), the objects of the `popim_population` class
also retain the input parameters of the basic constructor
`popim_population()`:

* `region`: a character vector of all regions in this object,
* `year_min` and `year_max`: the range of years, and
* `age_min` and `age_max`: the age range covered in this population.


#### Create, modify and validate objects of this class

create: `popim_population()`, `read_popim_pop()`, `as_popim_pop()`

modify: `apply_vacc()`

validate: `is_population()` *(currently not exported)*




### S3 class `popim_vacc_activities`

The class `popim_vacc_activities` is designed to hold information on
vaccination activities that have occurred or are planned in the
population of interest. Each row of the dataframe relates to one
vaccination activity.

#### The dataframe

The mandatory columns are:

* `region`: The region in which the vaccination activity takes
  place. While this is a free character field, when applied to a
  population, this must match a region that occurs in the population.
* `year`: year during which the vaccination activity takes place.
* `age_first`: youngest age group to be targeted.
* `age_last`: oldest age group to be targeted.
* `coverage`: proportion of the target population to be immunised.
* `doses`: number of doses used in the vaccination activity.
* `targeting`: determines how doses are allocated when there is
  pre-existing immunity in the population, see
  [below](#vaccinating-a-cohort-with-previous-immunity) for a
  description of the different `targeting` methods.

| column      | type      | range             |
|-------------|-----------|-------------------|
| `region`    | character |                   |
| `year`      | integer   |                   |
| `age_first` | integer   | non-negative      |
| `age_last`  | integer   | $\ge$ `age_first` |
| `coverage`  | numeric   | [0, 1]            |
| `doses`     | numeric   | non-negative      |
| `targeting` | character | `"random"`, `"correlated"`, `"targeted"` |

##### Relationship between coverage, doses and target population

If the target population size is known, the information on coverage
and doses is redundant (and potentially conflicting) as coverage =
doses / target population size. However, the `popim_vacc_activities`
object does not store information on population size, this is held in
the `popim_population` object to which the vaccination activities will
be applied, so this potential conflict cannot be detected in a
`popim_vacc_activities` object in isolation. The validator for this
class, `validate_vacc_activities()`, checks that the required columns
exist and are of correct data type and range (where applicatble). It
also currently requires that at least one of `coverage` and `doses` is
non-missing in each column. *(This behaviour is maybe not sensible as
inconsistent with the other columns?)*

#### Class attributes

As this is a subclass of dataframe, it has the same attributes as a
dataframe (`names`, `class` and `row_names`). It has no additional
attributes.

#### Create and validate objects of this class

create: `popim_vacc_activities()`, `read_vacc_activities()`,
`as_vacc_activities()`, `vacc_from_immunity()`

modify: `complete_vacc_activities()`

validate: `validate_vacc_activities()` *(currently not exported)*

## Primary functionality: applying vaccination activities to a population

```{r, results = "asis"}
child_env <- new.env()
child_env$type <- "vignette"

res <- knitr::knit_child("../man/rmd/apply.Rmd", envir = child_env, quiet = TRUE)

cat(res, sep = "\n")

```



### The inverse: inferring vaccination activities from a population

It is not possible to infer the vaccination activities that have given
rise to a particular immunity profile of a population by age and over
time: The same profile can be reached with different `targeting`
choices (where different `targeting` choices will require different
amount of vaccine). Furthermore, if there are cohorts that reached
100\% immunity, there may have been double vaccination of some
individuals that left no trace in the immunity profile. However, the
function `vacc_from_immunity()` does infer the vaccination activities
from a population immunity profile as far as possible, returning an
obejct of class `popim_vacc_activities`.

To this end, in addition to passing the `popim_population` object of
interest, the user needs to specify a `targeting` option (which will
be assumed for all vaccination activities that are identified). The
inferred vaccination activities will be the minimal activities
possible to give rise to the observed immunity profile given the
`targeting` option supplied.


## Population summary & visualisation

An object of class `popim_population` holds a somewhat complex dataset
recording population size, age distribution and immunity status
through time, and potentially disaggregated into various geographical
regions. To facilitate interpretation of this kind of data, there are
a couple of functions to aggregate and visualise these objects.

### Visualisation

The functions `plot_pop_size()` and `plot_immunity()` serve to plot
the age-disaggregated population size and immunity through time in a
grid where year and age are plotted on the x- and y-axis,
respectively. The size of the cohort or proportion of the cohort that
is immune is indicated by the colour of the grid cell. If there are
several regions, these will be shown in separate panels.

The implementation of these plot functions is based on ggplot2, and
therefore the returned plot objects can be further modified using the
ggplot2 syntax.

### Aggregation over age

While the age structure of a population's immunity profile is
important to understand how immunity is likely to develop through
time, for the purposes of understanding how well a population is
protected at any point in time, often the overall proportion immune is
of greater interest. This can be calculated using the function
`calc_pop_immunity()` which will aggregate the population over age and
return a dataframe with the whole population immunity over time and
geographical region.





