---
title: "Useful data sources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Useful data sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.dim = c(8,6)
)
```

```{r setup}
library(popim)
```
When using the the `popim` package it can be useful to initialise
populations and vaccination activities with realistic data.

## Population data

The age strucutred population size through time is a heavy data
requirement, but the 
[UN World Population Prospects](https://population.un.org/wpp/) has
population data in the right format for use with popim.

There is an API that can be accessed programmatically from R, but I
haven't managed to get this to work yet.

However, country-based population size by single-year age group can be
downloaded in .csv file format at the 
[Data Portal](https://population.un.org/dataportal/). 

The appropriate input data for `popim` are the "Population by one year
age groups and sex" for the countries of interest with the maximal
date range 1950 - 2101.

In these data, countries are labelled by name, but also by the ISO3
code. A list of the countries included is included in the package as
the object `unwpp_countries`:
```{r}
data(unwpp_countries)

dim(unwpp_countries)
head(unwpp_countries)

```

Here we focus on Nigeria (NGA) and India (IND) as an example, having
downloaded the relevant dataset in .csv format from the UNWPP Data
Portal. Here we have restricted the period to 2001 to 2030 for a more
moderate file size.


```{r}

pop <- read.csv("WPP2024_Population1JanuaryBySingleAgeSex_Medium_2001-2030_IND_NGA.csv",
                stringsAsFactors = FALSE, na.strings = "<NA>",
                encoding = "UTF-8")

my_countries <- c("India", "Nigeria")


library(dplyr)
drop_unwpp_cols <- function(dat, countries) {
    dat |> filter(Location %in% countries) |>
    select(Location, Time, AgeStart, Value) |>
    rename(region = Location, year = Time, age = AgeStart,
           pop_size = Value)
}

pop <- drop_unwpp_cols(pop, my_countries)
pop <- as_popim_pop(pop)

```

The plot of the absolute population size reveals that India's
population size dwarfs that of Nigeria, but obscures the age
distribution of Nigeria a little.

```{r}
plot_pop_size(pop)
```

The age distributions of both countries can be seen more clearly when
plotting relative population sizes:

```{r}
plot_pop_size(pop, rel = TRUE)
```

<!-- ## Vaccination data -->
